### 2024-05-31 manuscript/CD_submission1/abstract.qmd
cleaning up manuscript folder

---
bibliography: references.bib
csl: apa.csl
format:
  docx:
    reference-doc: reference-doc.docx
output:
  officedown::rdocx_document:
    page_margins:
      bottom: 1
      footer: 0
      gutter: 0
      header: 0.5
      left: 1
      right: 1
      top: 1
    plots:
      align: center
      caption:
        pre: 'Figure '
        sep: '. '
        style: Image Caption
    tables:
      caption:
        pre: 'Table '
        sep: '. '
        style: Table Caption
  pdf_document: default
  word_document: default
editor: 
  markdown: 
    wrap: sentence
---

>>>>

---
bibliography: ../references.bib
csl: apa.csl
format:
  docx:
    reference-doc: ../reference-doc.docx
output:
  officedown::rdocx_document:
    page_margins:
      bottom: 1
      footer: 0
      gutter: 0
      header: 0.5
      left: 1
      right: 1
      top: 1
    plots:
      align: center
      caption:
        pre: 'Figure '
        sep: '. '
        style: Image Caption
    tables:
      caption:
        pre: 'Table '
        sep: '. '
        style: Table Caption
  pdf_document: default
  word_document: default
editor: 
  markdown: 
    wrap: sentence
---

### 2024-05-31 manuscript/CD_submission1/manuscript.qmd
cleaning up manuscript folder

# Source necessary objects
power_linear      <- read_csv("../data/1_pilot/simulation/power_linear_model.csv") %>% mutate(power = round(power, 0))
power_mixed       <- read_csv("../data/1_pilot/simulation/power_mixed_model.csv") %>% mutate(power = round(power, 0))

load("pilot_staged_results.RData")
load("study1_staged_results.RData")
load("study2_staged_results.RData")
load("pooled_staged_results.Rdata")
```

@@

All scripts, materials, and data supporting the results presented here are available on the article's GitHub repository [ANONYMIZED FOR REVIEW].

@@

Thus, lowered performance on tasks like the Flanker could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2022].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

knitr::include_graphics("figures/fig1.png")
```

@@

In the Pilot, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks.
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
We expected violence exposure to be positively associated with a present-oriented attention style.
Specifically, we expected participants with more violence exposure to be better at detecting peripheral stimuli and subtle changes.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
Second, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).

@@

```{r}
#| label: figure2
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 2.** Multiverse results for the association between violence exposure with the strength of perceptual input and interference on the Flanker across all studies. Each panel depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). The top row depicts effect curves in the Pilot. The second row depicts effect curves in Study 1. The third row depicts effect curves in Study 2. The Fourth row depicts effect curves of the pooled analyses across all studies.
knitr::include_graphics("figures/fig2.png")
```

@@

```{r}
#| label: figure3
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 3.** Multiverse results for the association between unpredictability with the strength of perceptual input and interference on the Flanker across all studies. Each panel depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). The top row depicts effect curves in the Pilot. The second row depicts effect curves in Study 1. The third row depicts effect curves in Study 2. The Fourth row depicts effect curves of the pooled analyses across all studies.
knitr::include_graphics("figures/fig3.png")
```

@@

```{r}
#| label: figure4
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 4.** Multiverse results for the within-subjects comparison of Flanker and Global-Local information processing. Panel A depicts the multiverse interaction effects, with the thick black lines denoting the median slope and the thin lines denoting effects for each combination of arbitrary decisions. Blue thin lines indicate significant effects (*p* > .05), and grey thin lines indicate non-significant effects (*p* > .05). Panel B depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). See the main text for more information about the multiverse analyses.
knitr::include_graphics("figures/fig4.png")
```

>>>>

# Source necessary objects
power_linear      <- read_csv("../../data/1_pilot/simulation/power_linear_model.csv") %>% mutate(power = round(power, 0))
power_mixed       <- read_csv("../../data/1_pilot/simulation/power_mixed_model.csv") %>% mutate(power = round(power, 0))

load("../pilot_staged_results.RData")
load("../study1_staged_results.RData")
load("../study2_staged_results.RData")
load("../pooled_staged_results.Rdata")
```

@@

All scripts, materials, and data supporting the results presented here are available on the article's GitHub repository (<https://anonymous.4open.science/r/CogDev_AnonLink>).

@@

Thus, lowered performance on tasks like the Flanker could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@fields_2021; @mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2022].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

knitr::include_graphics("../figures/fig1.png")
```

@@

In the Pilot, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks.
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
We expected people with more violence exposure to be better at detecting peripheral stimuli and subtle changes, as a result of a more diffuse scope of attention.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
Second, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).

@@

```{r}
#| label: figure2
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 2.** Multiverse results for the association between violence exposure with the strength of perceptual input and interference on the Flanker across all studies. Each panel depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). The top row depicts effect curves in the Pilot. The second row depicts effect curves in Study 1. The third row depicts effect curves in Study 2. The Fourth row depicts effect curves of the pooled analyses across all studies.
knitr::include_graphics("../figures/fig2.png")
```

@@

```{r}
#| label: figure3
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 3.** Multiverse results for the association between unpredictability with the strength of perceptual input and interference on the Flanker across all studies. Each panel depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). The top row depicts effect curves in the Pilot. The second row depicts effect curves in Study 1. The third row depicts effect curves in Study 2. The Fourth row depicts effect curves of the pooled analyses across all studies.
knitr::include_graphics("../figures/fig3.png")
```

@@

```{r}
#| label: figure4
#| fig.width: 7
#| fig.height: 9
#| dpi: 600
#| fig-cap: | 
#|   **Figure 4.** Multiverse results for the within-subjects comparison of Flanker and Global-Local information processing. Panel A depicts the multiverse interaction effects, with the thick black lines denoting the median slope and the thin lines denoting effects for each combination of arbitrary decisions. Blue thin lines indicate significant effects (*p* > .05), and grey thin lines indicate non-significant effects (*p* > .05). Panel B depicts sorted beta coefficients across all combinations of arbitrary decisions (i.e., the effect curve across the whole multiverse). See the main text for more information about the multiverse analyses.
knitr::include_graphics("../figures/fig4.png")
```

### 2024-05-31 manuscript/CD_submission1/manuscript_blinded.qmd
cleaning up manuscript folder

#### **Childhood adversity is not associated with lowered inhibition, but slower perceptual processing: A Drift Diffusion Model analysis**

#### Stefan Vermeent^a,b^, Ethan S. Young^a^, Jean-Louis van Gelder^b^, & Willem E. Frankenhuis^a,b,c^

#### ^a^ Department of Psychology, Utrecht University, Utrecht, The Netherlands

#### ^b^ Max Planck Institute for the Study of Crime, Security, and Law, Freiburg, Germany

#### ^c^ Evolutionary and Population Biology, Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Amsterdam, the Netherlands

# Corresponding author

Correspondence concerning this article should be addressed to Stefan Vermeent, Department of Psychology, Utrecht University, Heidelberglaan 1, 3584 CS Utrecht, The Netherlands. Email: p.c.s.vermeent@gmail.com / p.c.s.vermeent@uu.nl.

# Data Statement

All scripts, materials, and data supporting the results presented here are available on the article's GitHub repository (<https://anonymous.4open.science/r/CogDev_AnonLink>).

# Funding Statement
WEF’s contributions have been supported by the Dutch Research Council (V1.Vidi.195.130) and the James S. McDonnell Foundation (https://doi.org/10.37717/220020502).

# Author contributions

**Stefan Vermeent:** Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Project administration, Data curation, Visualization, Writing - original draft, Writing - review & editing.
**Ethan S. Young:** Conceptualization, Methodology, Supervision, Writing - review & editing.
**Jean-Louis van Gelder:** Conceptualization, Funding acquisition, Methodology, Supervision, Writing - review & editing.
**Willem E. Frankenhuis:** Conceptualization, Funding acquisition, Methodology, Supervision, Writing - review & editing.

# Abstract

It is well-established that individuals who grew up in adverse conditions tend to be slower on the Flanker Task. 
This finding is typically interpreted to reflect difficulty inhibiting distractions. 
However, it might result from slower general cognitive processes (e.g., reduced general processing speed), rather than the specific ability of inhibition. 
We used Drift Diffusion Modeling in three online studies (total N = 1560) with young adults to understand associations of adversity with Flanker performance. 
We find no associations between exposure to violence and unpredictability with inhibition. 
Yet, although mixed, violence and unpredictability exposure were associated with lower strength of perceptual input--—how well someone can process target and distractor information alike. 
Finally, people with lower strength of perceptual input processed information more holistically, focusing less on details. 
Thus, lowered Flanker performance does not necessarily imply lowered inhibition ability. 
Cognitive modeling might reveal a different picture of abilities in adverse conditions.

*Key-words:* childhood adversity, inhibition, Flanker, Drift Diffusion Modeling

@@

Thus, lowered performance on tasks like the Flanker could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@fields_2021; @mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2022].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

In the Pilot, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks.
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
We expected people with more violence exposure to be better at detecting peripheral stimuli and subtle changes, as a result of a more diffuse scope of attention.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
Second, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).

>>>>

@@

Thus, lowered performance on tasks like the Flanker could reflect either a cognitive impairment or a difference in attentional strategies.
Distinguishing between these two possibilities is challenging for two reasons.
First, few studies in the adversity literature have measured performance differences on different attention tasks within the same individual [@mezzacappa_2004; @mittal_2015].
Thus, it is unclear whether lowered inhibition is related to enhanced processing of peripheral information in people from from adverse backgrounds.
Second, performance on inhibition tasks is---beyond the ability to inhibit distractors---also influenced by other factors, such as a person's general processing speed and response caution [@hedge_2022; @loffler_2022].
This means that lowered performance on inhibition tasks does not necessarily reflect inhibition difficulties.
In other words, we should consider cognitive processes other than ability when drawing inferences based on inhibition tasks.

@@

In the Pilot, our goal was to understand how childhood adversity relates to performance on tasks with different attentional demands.
Participants completed self-report measures of childhood adversity and three cognitive tasks.
These tasks measured inhibition, attention for peripheral cues, and attention for subtle changes.
We expected violence exposure to be positively associated with a present-oriented attention style.
Specifically, we expected participants with more violence exposure to be better at detecting peripheral stimuli and subtle changes.
We expected this would result in a higher drift rate (faster speed of information accumulation) or shorter non-decision times (faster attention orientation, among other things), but not necessarily with differences in boundary separation (response caution).
Second, we expected that participants with more violence exposure would be worse at ignoring distracting peripheral stimuli.
We expected this would result in more experienced interference (as derived from the SSP model).
